### 2025-07-13 docs/article/Sections/state-of-the-art-web-templates.tex
Article section 2: Update template engines table type-safety section; added Qute.
Section 3 & 5: update template engines table; Add HtmlFlow-Rx to Quarkus Results and change Sync to Blocking
Post MDPI review changes.

@@

@@

     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher}$^{(1)}$}
     & \texttt{Publisher}$^{(1)}$

@@

     & \shortstack{\texttt{Iterable} \\\texttt{Sequence}$^{(2)}$}

@@

     & \textcolor{PineGreen}{\checkmark}$^{(3)}$

@@

    \\$^{(1)}$ \texttt{Publisher} of Reactive Streams~\cite{ReactiveStreams}.
    Limited to a single model per web template.
    \\$^{(2)}$ \texttt{kotlin.sequences.Sequence}
    \\$^{(3)}$ Non-safety for HTML attributes.

@@

Web templates are based on a \textit{domain-specific language}
(DSL)~\cite{landin1966next}, which defines a language tailored to a specific
\textit{domain}~\cite{evans2004domain}—in this case, HTML for expressing web
documents. The DSL constrains the template's syntax and semantics to match the
structure and purpose of HTML.
DSLs can be divided in two types: \textit{external} or
\textit{internal}\cite{dslbook}. \textit{External} DSLs are languages created
without any affiliation to a concrete programming language. An example of an
\textit{external} DSL is the regular expressions search
pattern\cite{thompson1968}, since it defines its own syntax without any
dependency of programming languages. On the other hand an \textit{internal} DSL
is defined within a host programming language as a library and tends to be
limited to the syntax of the host language, such as Java.
JQuery\cite{resig2007pro} is one of the most well-known examples of an internal
DSL in Javascript, designed to simplify HTML DOM\cite{dom} tree traversal and
manipulation.

@@

From the three examples depicted in Figure~\ref{fig:dsl-idioms}, the
\textit{nested function} idiom used by \textbf{ScalaTags}, as shown in the
snippet of Figure~\ref{fig:dsl-idioms}.b), is the least verbose, requiring
fewer statements than the other two cases. This approach of combining function
calls is also utilized by \textbf{Hiccup} Clojure Library and \textbf{j2html}
Java library. One reason for verbosity in \textit{function sequence} and
\textit{method chaining} approaches is the requirement of a dedicated HTML
builder to emit the closing tag, exemplified by \texttt{end()} in \textbf{JATL}
(Figure~\ref{fig:dsl-idioms}.a) and \texttt{\_\_()} in \textbf{HtmlFlow}
(Figure~\ref{fig:dsl-idioms}.c). But the \textit{nested function} approach
comes with a notable drawback: \textbf{they do not support PSSR} because the
sequence of nested functions is evaluated backward to the order in which they
are written. In other words, arguments are evaluated before the functions are
invoked. Taking the example in Figure~\ref{fig:dsl-idioms}.b), the
\texttt{title()} function is first evaluated, and its resulting paragraph
becomes the argument for the \texttt{head()} call, which, in turn, becomes the
argument for \texttt{html()}, and so on. If HTML is emitted as functions are
called, it will print tags in reverse order. The aforementioned DSLs must
collect resulting nodes into an internal data structure, which is later
traversed to produce the HTML output. Therefore, they cannot progressively emit
the output as builders are called.

@@

To achieve this KotlinX.html provides HTML builders using \textit{function
  literals with receiver}~\cite{kotlinlang}. In Kotlin, a block of code enclosed
in curly braces \texttt{\{...\}} is known as a \emph{lambda}, and can be used
as an argument to a function that expects a \emph{function literal}. When we
write, for example, \texttt{body \{ div \{ hr() \} \}}, we are invoking the
\texttt{body} function with a lambda as its argument. This lambda, in turn,
calls the \texttt{div} function with another lambda as an argument that creates
a horizontal row (i.e. \texttt{hr}). Each call to an HTML builder (e.g.,
\texttt{body}, \texttt{div}, \texttt{hr}) creates the child element within the
element generated by the outer function call.

@@

\textit{\textbf{Progressive rendering}} and \textit{\textbf{progressive loading}}
encompass different concepts. The former pertains to the \textit{dynamic} content of a
\textit{dynamic web page}, encompassing elements with logic and placeholders
that are fulfilled by data from an object \textit{model} constructed at runtime.
On the other hand, the latter is associated with \textit{render-blocking
  resources} such as scripts, stylesheets, and HTML imports, which may hinder the
browser from rendering page content to the screen~\cite{progressive-2022}.
The notion of \textit{progressive rendering} has typically been aligned with
\textit{client-side rendering} (CSR), where a single HTML page with static
content is delivered upfront, while the dynamic content is fetched as the data
becomes available to complete the web page.
However, despite being overlooked, HTTP and browsers were designed from their
inception to also support this feature in the context of SSR approaches,
which offers the advantage of not being dependent on
\textit{rendering-blocking resources}, such as the required JavaScript for CSR.
% \textit{Progressive server-side rendering} (PSSR) is the
% ability of streaming HTML content to the client in chunks or data frames, as it
% gets resolved on the server at runtime.
% This lets the user-interface to be rendered incrementally (i.e.
% \textit{progressively}) by the browser in line with the availability of the data.
% However, most SSR template engines do not inherently support this feature at the
% template level, requiring developers to manually break down web pages into
% fragments or partials to enable PSSR.
An example of this limitation in SSR web templates was highlighted by Jeff
Atwood in 2005, who criticized Microsoft ASP.NET for loading the entire web
page into memory before sending any data to the browser~\cite{pssr2005}.
% This practice delays the display of content to the user until the server
% completes its processing, contradicting the original design intent of HTML to
% render progressively as content is received.
% Notably, both Netscape and Internet Explorer were capable of rendering partial
% HTML content from their inception.
Despite historical critiques and HTML's inherent capabilities, most Web
application frameworks, such as ASP.Net, Express.js, Spring, and others,
persistently lack support for progressive rendering, leading to the appearance
of alternative techniques leveraging client-side JavaScript. Since 2007,
various patents have addressed the PSSR issue, with Microsoft's patent enabling
the infinite scrolling technique by displaying a single page of
results~\cite{scroll2007}, and Yahoo's patent focusing on differentiating
elements based on their position relative to the visible
portion~\cite{schiller2007progressive}.

>>>>

     & \textcolor{PineGreen}{\checkmark}$^{(1)}$

@@

    \textbf{Qute}
     & \textcolor{PineGreen}{\checkmark}$^{(2)}$

@@

     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher}$^{(3)}$}
     & \texttt{Publisher}$^{(3)}$

@@

     & \shortstack{\texttt{Iterable} \\\texttt{Sequence}$^{(4)}$}

@@

     & \textcolor{PineGreen}{\checkmark}$^{(5)}$

@@

    \\$^{(1)}$ Enforced via annotation processor that checks templates against typed interfaces at compile time.
    \\$^{(2)}$ Compile-time expression validation available via \texttt{@CheckedTemplate} and build-time metadata in Quarkus.
    \\$^{(3)}$ \texttt{Publisher} of Reactive Streams~\cite{ReactiveStreams}. Limited to a single model per web template.
    \\$^{(4)}$ \texttt{kotlin.sequences.Sequence}
    \\$^{(5)}$ Non-safety for HTML attributes.

@@

Web templates are based on a \textit{domain-specific language}
(DSL)~\cite{landin1966next}, which defines a language tailored to a specific
\textit{domain}~\cite{evans2004domain}—in this case, HTML for expressing web
documents. The DSL constrains the template's syntax and semantics to match the
structure and purpose of HTML.
DSLs can be divided in two types: \textit{external} or
\textit{internal}\cite{dslbook}. \textit{External} DSLs are languages created
without any affiliation to a concrete programming language. An example of an
\textit{external} DSL is the regular expressions search
pattern\cite{thompson1968}, since it defines its own syntax without any
dependency of programming languages. On the other hand, an \textit{internal} DSL
is defined within a host programming language as a library and tends to be
limited to the syntax of the host language, such as Java.
JQuery\cite{resig2007pro} is one of the most well-known examples of an internal
DSL in Javascript, designed to simplify HTML DOM\cite{dom} tree traversal and
manipulation.

@@

From the three examples depicted in Figure~\ref{fig:dsl-idioms}, the
\textit{nested function} idiom used by \textbf{ScalaTags}, as shown in the
snippet of Figure~\ref{fig:dsl-idioms}.b), is the least verbose, requiring
fewer statements than the other two cases. This approach of combining function
calls is also utilized by \textbf{Hiccup} Clojure Library and \textbf{j2html}
Java library. One reason for verbosity in \textit{function sequence} and
\textit{method chaining} approaches is the requirement of a dedicated HTML
builder to emit the closing tag, exemplified by \texttt{end()} in \textbf{JATL}
(Figure~\ref{fig:dsl-idioms}.a) and \texttt{\_\_()} in \textbf{HtmlFlow}
(Figure~\ref{fig:dsl-idioms}.c). But the \textit{nested function} approach
comes with a notable drawback: \textbf{it does not support PSSR} because the
sequence of nested functions is evaluated backward to the order in which they
are written. In other words, arguments are evaluated before the functions are
invoked. Taking the example in Figure~\ref{fig:dsl-idioms}.b), the
\texttt{title()} function is first evaluated, and its resulting paragraph
becomes the argument for the \texttt{head()} call, which, in turn, becomes the
argument for \texttt{html()}, and so on. If HTML is emitted as functions are
called, it will print tags in reverse order. The aforementioned DSLs must
collect resulting nodes into an internal data structure, which is later
traversed to produce the HTML output. Therefore, they cannot progressively emit
the output as builders are called.

@@

To achieve this, KotlinX.html provides HTML builders using \textit{function
  literals with receiver}~\cite{kotlinlang}. In Kotlin, a block of code enclosed
in curly braces \texttt{\{...\}} is known as a \emph{lambda}, and can be used
as an argument to a function that expects a \emph{function literal}. When we
write, for example, \texttt{body \{ div \{ hr() \} \}}, we are invoking the
\texttt{body} function with a lambda as its argument. This lambda, in turn,
calls the \texttt{div} function with another lambda as an argument that creates
a horizontal row (i.e. \texttt{hr}). Each call to an HTML builder (e.g.,
\texttt{body}, \texttt{div}, \texttt{hr}) creates the child element within the
element generated by the outer function call.

@@

\textit{\textbf{Progressive rendering}} and \textit{\textbf{progressive loading}}
encompass different concepts. The former pertains to the \textit{dynamic} content of a
\textit{dynamic web page}, encompassing elements with logic and placeholders
that are fulfilled by data from an object \textit{model} constructed at runtime.
On the other hand, the latter is associated with \textit{render-blocking
  resources} such as scripts, stylesheets, and HTML imports, which may hinder the
browser from rendering page content to the screen~\cite{progressive-2022}.
The notion of \textit{progressive rendering} has typically been aligned with
\textit{client-side rendering} (CSR), where a single HTML page with static
content is delivered upfront, while the dynamic content is fetched as the data
becomes available to complete the web page.
However, despite being overlooked, HTTP and browsers were designed from their
inception to also support this feature in the context of SSR approaches,
which offers the advantage of not being dependent on
\textit{render-blocking resources}, such as the required JavaScript for CSR.
% \textit{Progressive server-side rendering} (PSSR) is the
% ability of streaming HTML content to the client in chunks or data frames, as it
% gets resolved on the server at runtime.
% This lets the user-interface to be rendered incrementally (i.e.
% \textit{progressively}) by the browser in line with the availability of the data.
% However, most SSR template engines do not inherently support this feature at the
% template level, requiring developers to manually break down web pages into
% fragments or partials to enable PSSR.
An example of this limitation in SSR web templates was highlighted by Jeff
Atwood in 2005, who criticized Microsoft ASP.NET for loading the entire web
page into memory before sending any data to the browser~\cite{pssr2005}.
% This practice delays the display of content to the user until the server
% completes its processing, contradicting the original design intent of HTML to
% render progressively as content is received.
% Notably, both Netscape and Internet Explorer were capable of rendering partial
% HTML content from their inception.
Despite historical critiques and HTML's inherent capabilities, most Web
application frameworks, such as ASP.Net, Express.js, Spring, and others,
persistently lack support for progressive rendering, leading to the appearance
of alternative techniques leveraging client-side JavaScript. Since 2007,
various patents have addressed the PSSR issue, with Microsoft's patent enabling
the infinite scrolling technique by displaying a single page of
results~\cite{scroll2007}, and Yahoo's patent focusing on differentiating
elements based on their position relative to the visible
portion~\cite{schiller2007progressive}.
